---
title: "Generalized Linear models"
author: "Mark Andrews"
date: "April 5, 2017"
output:
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

# Introduction

In generalized linear models, we model the outcome variable as a random variable whose parameters are transformed linear functions of some of more predictors variables. 

```{r message=FALSE}
library(pander)
```


# Logistic regression

In a binary logistic regression, we model the outcome variable as Bernoulli random variable with a parameter $p$, and where the log odds of $p$ is a linear function of predictor variables. In other words, for all $i$,
$$
\begin{aligned}
  y_i &\sim \textrm{dbern}(p_i), \\
  \log\left(\frac{p_i}{1-p_i}\right) &= \beta_0 + \sum_{k=1}^K \beta_k x_{ki}  
\end{aligned}
$$

We'll load up some data about extra marital affairs. 
```{r}
load('../data/affairs.Rda')
```

Let's take a look:
```{r}
head(Affairs)
```
We create a new variable that indicates if someone cheats or not:
```{r}
Affairs$cheater <- Affairs$affairs > 0
```

Now, we'll model how the probability of cheating varies by gender:
```{r}
M <- glm(cheater ~ gender, 
         data=Affairs, 
         family=binomial)
```

## Predictions
As usual, we will make some data to make predictions about:
```{r}
hypothetical.data <- data.frame(gender=c('male', 'female'))
```
and then make the predictions
```{r}
predict(M, newdata=hypothetical.data)
```
These predictions are in log odds units, so we can convert to probabilities using the inverse logit function, which we can make ourselves:
```{r}
ilogit <- function(x){1/(1+exp(-x))}

logodds <- predict(M, newdata=hypothetical.data) # these are log odds
names(logodds) <- c('Male', 'Female')
ilogit(logodds)
```

We can get the same result more easily with the following:
```{r}
predictions <- predict(M, newdata=hypothetical.data, type='response') 
names(predictions) <- c('Male', 'Female')
predictions
```

## Model comparison

We will model cheating using two different models, i.e. two models with different numbers of predictors:

```{r}
### Using all predictors
M <- glm(cheater ~ gender + age + yearsmarried
         + children + religiousness + education
         + occupation + rating, 
         data=Affairs, 
         family=binomial)

# This is the "null" model, i.e. no predictors
M.null <- glm(cheater ~ 1, 
              data=Affairs, 
              family=binomial)
```

We do model comparison by way of a log likelihood test:
```{r}
ll.test <- anova(M.null, M, test='Chisq')
pander(ll.test, missing='')
```


# Poisson regression

Instead of modelling the probability of cheating, we can model the number of affairs people have, using a Poisson regression model:

```{r}
M <- glm(affairs ~ gender + age + yearsmarried
         + children + religiousness + education
         + occupation + rating, 
         data=Affairs, 
         family=poisson)

M.null <- glm(affairs ~ 1, 
              data=Affairs, 
              family=poisson)

# Model fit comparison of null and full based on the "Deviance"
ll.test <- anova(M.null, M, test='Chisq')
pander(ll.test, missing='')
```





