---
title: "Multilevel models"
author: "Mark Andrews"
date: "November 14, 2018"
output:
  pdf_document: default
editor_options:
  chunk_output_type: inline
---


```{r, message=FALSE}
library(lme4)
library(lmerTest)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
```

Let's consider the following data:
```{r}
ggplot(sleepstudy,
       aes(x=Days, y=Reaction, col=Subject)
) + geom_point() +
  stat_smooth(method='lm', se=F) +
  facet_wrap(~Subject) +
  guides(col=F)
```

If we label our the reaction time, subject, and day on observation $i$ by $y_i$, $s_i \in \{1, 2 \ldots J\}$, and $x_i$, respectively, a random intercepts model of this data would be
$$
\begin{aligned}
y_i \sim N(a_{s_i} + b x_i, \sigma^2),\quad\text{for all $i \in 1, 2\ldots n$}\\
a_j \sim N(\alpha, \tau^2),\quad\text{for all $j \in 1, 2\ldots J$}
\end{aligned}
$$


```{r}
M_0 <- lmer(Reaction ~ Days + (1|Subject),
            data = sleepstudy)
```

A random slopes model of this data would be
$$
\begin{aligned}
y_i \sim N(a + b_{s_i} x_i, \sigma^2),\quad\text{for all $i \in 1, 2\ldots n$}\\
b_j \sim N(\beta, \tau_{\beta}^2),\quad\text{for all $j \in 1, 2\ldots J$}
\end{aligned}
$$


```{r}
M_1 <- lmer(Reaction ~ Days + (0 + Days|Subject),
            data = sleepstudy)
```

A random slopes and random intercepts model would be
$$
\begin{aligned}
y_i \sim N(a_{s_i} + b_{s_i} x_i, \sigma^2),\quad\text{for all $i \in 1, 2\ldots n$}\\
a_j \sim N(\alpha, \tau_{\alpha}^2),\quad\text{for all $j \in 1, 2\ldots J$}\\
b_j \sim N(\beta, \tau_{\beta}^2),\quad\text{for all $j \in 1, 2\ldots J$}
\end{aligned}
$$


```{r}
M_1 <- lmer(Reaction ~ Days + (1 + Days|Subject),
            data = sleepstudy)
```

These models can be re-written using different, but equivalent, notation. For example, the random slopes, random intercepts models is equivalent to the following:
$$
\begin{aligned}
y_i = \alpha + \beta x_i + \nu_{s_i} + \psi_{s_i} x_i + \epsilon_i,\quad\text{for all $i \in 1, 2\ldots n$}\\
\nu_j \sim N(0, \tau_{\alpha}^2),\quad\text{for all $j \in 1, 2\ldots J$}\\
\psi_j \sim N(0, \tau_{\beta}^2),\quad\text{for all $j \in 1, 2\ldots J$}.
\end{aligned}
$$
In this, $\nu_j = a_j - \alpha$ and $\psi_j = b_j - \beta$.

# Nested models

Sometimes we have groups nested in other groups
```{r, message=FALSE, warning=FALSE}
Df <- read_csv('../data/science.csv')
```

In `Df`, we have `class`, with values $\{1, 2, 3, 4\}$, nested in `school`, with values $\{1, 2 \ldots 41\}$. For example,
```{r}
group_by(Df, school, class) %>% 
  summarise(n = n()) %>% 
  spread(class,n, fill=0)
```

To model this nesting, we'd do the following:
```{r}
M_1 <- lmer(like ~ sex + PrivPub + (1|school/class), 
          data = Df)
```
which is identical to
```{r}
M_2 <- lmer(like ~ sex + PrivPub + (1|school) + (1|school:class), 
          data = Df)
```

However, if we use unique identifiers for `class`, i.e. `Class`, which takes values `1.1`, `1.2`, etc., then we can simply do
```{r}
M_3 <- lmer(like ~ sex + PrivPub + (1|school) + (1|Class), 
          data = Df)
```

# Crossed structures

When grouping variables are not nested, they are *crossed*. For example, in a biochemistry experiment, we could have six different samples of penicillian and 24 different plates. If each sample occurs in each plate, then we have a *fully crossed* structure. But fully crossed structures are not necessary.

```{r}
M <- lmer(diameter ~ 1 + (1|plate) + (1|sample),
          data=Penicillin)
```

# Model comparison

We proceed just like in the case of generalized linear models.
```{r}
M_null <- lmer(diameter ~ 1 + (1|sample),
               data=Penicillin)

anova(M_null, M)

```

