---
title: "Simple Statistical Data Analysis"
author: "Mark Andrews"
date: "April 5, 2017"
output:
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

# Introduction

With extremely rare exceptions, any statistical method that you might like to use for data analysis is probably already available in R. Here, we will start with the basics.

# T-tests

```{r}
set.seed(101)
x <- rnorm(10)
y <- rnorm(12)

t.test(x, y)
```

If we want to assume that the variance of the two groups are equal, then specify this as follows:
```{r}
M <- t.test(x, y, var.equal=T)
```

From this, the object *M* contains all the properties of the t-test results. For example, 
```{r}
M$statistic # The test statistic

M$parameter # The degrees of freedom

M$p.value # The p-value
```

## A paired sample t-test 
```{r}
set.seed(102)
N <- 10
x <- rnorm(N)
y <- rnorm(N)
(M <- t.test(x, y, paired = T))
```

## A one sample t-test 
```{r}
set.seed(103)
x <- rnorm(10)
(M <- t.test(x))
```

# Some non-parametric tests

Non-parametric counterparts of the independent samples and the paired samples t-tests are the Mann-Whitney U test and the Wilcoxon signed ranks tests.

This is the Mann Whitney U test:

```{r}
set.seed(101)

x <- rnorm(10)
y <- rnorm(12)

wilcox.test(x, y) # Yes, I know it is not called Mann Whitney
```

This is the Wilcoxon signed ranks test:
```{r}
set.seed(102)
N <- 10
x <- rnorm(N)
y <- rnorm(N)
wilcox.test(x, y, paired = TRUE) 
```

# Pearson's $\chi^2$ test

For this, we will use the *Titanic*[^titanic] data set. 
```{r}
data("Titanic") # load it up
```
This is a four dimensional table of frequencies:
```{r}
dimnames(Titanic)
```
We'll concatenate by 'Sex' and 'Survived' to make a 2 by 2 table to use as our observed frequencies:
```{r}
(observed <- apply(Titanic, c('Sex', 'Survived'), sum))
```

To do the $\chi^2$ test, it is simply
```{r}
(M <- chisq.test(observed))
```
As before, we can access properties of the test, e.g.
```{r}
M$expected
```

# Correlations
```{r}
set.seed(104)
N <- 20

x <- rnorm(N)
y <- rnorm(N)
```

To do a good old Pearson's product moment correlation:
```{r}
cor.test(x, y)
```
And a good old Spearman's $\rho$:
```{r}
cor.test(x, y, method='spearman')
```


[^titanic]: *Titanic* is a ship named after a famous movie from the 1990's.